---
title: "Functions and for loops"
---

## Motivation

- You have learned how to use different packages within the `tidyverse`, and two crucial packages for spatial data (`sf` and `terra`)
- These should help you work on ~80% of your use cases
- What about the other 20%?
- Other packages may help
- `do_science()` doesn't exist
- You need to be able to build your own code, from the ground up

## Two crucial things to learn

1. How to wrap multiple lines of code into a single function call
2. How to repeat the same calculation (perhaps with small tweaks) many times

. . .

> The goal: To do this with as few keystrokes as possible, and in a reproducible way

# Functions

## Functions

## When Might We Want to Build One?

. . .

> Whenever you’ve copied and pasted a block of code more than twice (i.e. you now have three copies of the same code).
>
> -- @Wickham2023-aa

. . .




When the function you want doesn't exist

## The General Approach

1. Analyze the code you are repeating. This will be your `function body`
2. Identify what parts are **constant** and what parts _vary_. The ones that vary will be `arguments`

. . .

This will allow you to have the basic architecture for your function

## An Example {.scrollable}

:::: {.columns}

::: {.column width='50%'}


```{r}
#| echo: false
# Load packages
library(tidyverse)
library(EVR628tools)
# Load data
data("data_lionfish")
```


```{r}
#| fig-height: 2
#| fig-width: 3.5
ggplot(data = data_lionfish,
       mapping = aes(x = total_length_mm,
                     y = total_weight_gr)) +
  geom_point() +
  labs(x = "Total length (mm)",
       y = "Total weight (gr)") +
  theme(axis.line = element_line(color = "black",
                                 linewidth = 0.5),
        axis.ticks = element_line(color = "black", 
                                  linewidth = 0.5),
        panel.grid.major = element_line(color = "black",
                                        linewidth = 0.5),
        panel.grid.minor = element_line(color = "black",
                                        linewidth = 0.5),
        panel.background = element_rect(fill = "lightblue"),
        axis.text = element_text(color = "black"),
        text = element_text(color = "black",
                            face = "bold",
                            size = 12))

ggplot(data = data_lionfish,
       mapping = aes(x = total_length_mm)) +
  geom_histogram() +
  labs(x = "Total Length (mm)",
       y = "N") +
  theme(axis.line = element_line(color = "black",
                                 linewidth = 0.5),
        axis.ticks = element_line(color = "black", 
                                  linewidth = 0.5),
        panel.grid.major = element_line(color = "black",
                                        linewidth = 0.5),
        panel.grid.minor = element_line(color = "black",
                                        linewidth = 0.5),
        panel.background = element_rect(fill = "lightblue"),
        axis.text = element_text(color = "black"),
        text = element_text(color = "black",
                            face = "bold",
                            size = 14))
```


:::

::: {.column width='50%'}

```{r}
#| fig-height: 2
#| fig-width: 3.5
ggplot(data = data_lionfish,
       mapping = aes(x = depth_m,
                     y = total_length_mm)) +
  geom_smooth(method = "lm") +
  labs(x = "Depth (m)",
       y = "Total Length (mm)") +
  theme(axis.line = element_line(color = "black",
                                 linewidth = 0.5),
        axis.ticks = element_line(color = "black", 
                                  linewidth = 0.5),
        panel.grid.major = element_line(color = "black",
                                        linewidth = 0.5),
        panel.grid.minor = element_line(color = "black",
                                        linewidth = 0.5),
        panel.background = element_rect(fill = "lightblue"),
        axis.text = element_text(color = "black"),
        text = element_text(color = "black",
                            face = "bold",
                            size = 10))

ggplot(data = data_lionfish,
       mapping = aes(x = depth_m,
                     y = total_weight_gr)) +
  geom_bin_2d() +
  labs(x = "Depth (mm)",
       y = "Total weight (gr)") +
  theme(axis.line = element_line(color = "black",
                                 linewidth = 0.5),
        axis.ticks = element_line(color = "black", 
                                  linewidth = 0.5),
        panel.grid.major = element_line(color = "black",
                                        linewidth = 0.5),
        panel.grid.minor = element_line(color = "black",
                                        linewidth = 0.5),
        panel.background = element_rect(fill = "lightblue"),
        axis.text = element_text(color = "black"),
        text = element_text(color = "black",
                            face = "bold",
                            size = 14))
```

:::

::::


## An Example

- You spent 10 hours building these and other 10 "beautiful" figures

- And now you are asked to modify the theme to use thinner lines and a light gray background

- **Q:** How many changes do you have to make to your code?

- **A:** 2 changes for each plot, with 12 plots = chaos

- But what if you had a function that describes your own theme?

- Then you would only need to make changes in 1 location

## Parts of a Function 

A function has three components:

::: {.nonincremental}

1. A name
2. Arguments
3. Body

:::

```{r}
#| code-fold: false
#| code-line-numbers: "1|2|"
name <- function(arguments){ # What changes
  body # What your function does
}
```

## Let's build our own theme

Look at the code again

:::: {.columns}

::: {.column width='50%'}
```{r}
#| eval: false
#| code-fold: false
ggplot(data = data_lionfish,
       mapping = aes(x = total_length_mm,
                     y = total_weight_gr)) +
  geom_point() +
  labs(x = "Total length (mm)",
       y = "Total weight (gr)") +
  theme(axis.line = element_line(color = "black",
                                 linewidth = 0.5),
        axis.ticks = element_line(color = "black", 
                                  linewidth = 0.5),
        panel.grid.major = element_line(color = "black",
                                        linewidth = 0.5),
        panel.grid.minor = element_line(color = "black",
                                        linewidth = 0.5),
        panel.background = element_rect(fill = "lightblue"),
        axis.text = element_text(color = "black"),
        text = element_text(color = "black",
                            face = "bold",
                            size = 12))
```

:::

::: {.column width='50%'}
```{r}
#| eval: false
#| code-fold: false
ggplot(data = data_lionfish,
       mapping = aes(x = depth_m,
                     y = total_length_mm)) +
  geom_smooth(method = "lm") +
  labs(x = "Depth (m)",
       y = "Total Length (mm)") +
  theme(axis.line = element_line(color = "black",
                                 linewidth = 0.5),
        axis.ticks = element_line(color = "black", 
                                  linewidth = 0.5),
        panel.grid.major = element_line(color = "black",
                                        linewidth = 0.5),
        panel.grid.minor = element_line(color = "black",
                                        linewidth = 0.5),
        panel.background = element_rect(fill = "lightblue"),
        axis.text = element_text(color = "black"),
        text = element_text(color = "black",
                            face = "bold",
                            size = 10)) # Font size changes to accomodate text
```

:::

::::

- Step 1: What are we repeating?
- Step 2: Within that code, what is constant and what changes?

## Test

## Build a function {.scrollable}

```{r}
#| code-fold: false
# Function name
my_theme <- function(text_size = 12) { # Function arguments
  # Function body
  theme(axis.line = element_line(color = "black",
                                   linewidth = 0.1),
          axis.ticks = element_line(color = "black", 
                                    linewidth = 0.1),
          panel.grid.major = element_line(color = "black",
                                          linewidth = 0.1),
          panel.grid.minor = element_line(color = "black",
                                          linewidth = 0.1),
          panel.background = element_rect(fill = "gray99"),
          axis.text = element_text(color = "black"),
          text = element_text(color = "black",
                              face = "bold",
                              size = text_size)) # Note how my argument isbeing used down here
}

```


## We Can Now Use Our Function

:::: {.columns}

::: {.column width='50%'}
```{r}
#| fig-height: 2
#| fig-width: 3.5
#| code-fold: false
#| code-line-numbers: "|7"
ggplot(data = data_lionfish,
       mapping = aes(x = total_length_mm,
                     y = total_weight_gr)) +
  geom_point() +
  labs(x = "Total length (mm)",
       y = "Total weight (gr)") +
  my_theme(text_size = 10)
```

:::

::: {.column width='50%'}
```{r}
#| fig-height: 2
#| fig-width: 3.5
#| code-fold: false
#| code-line-numbers: "|7"
ggplot(data = data_lionfish,
       mapping = aes(x = depth_m,
                     y = total_length_mm)) +
  geom_smooth(method = "lm") +
  labs(x = "Depth (m)",
       y = "Total Length (mm)") +
  my_theme(text_size = 8)
```

:::

::::

## Another example

- Functions are incredibly useful for computer simulations
- Imagine you want to simulate the growth of a population
- The simplest mathematical model of density-dependent growth is:

. . .

$$N_{t+1} = N_t + \left(rN_t\left(1 - \frac{N_t}{K}\right)\right)$$

. . .

> Read this as: The population size $N$ at time $t+1$ is the population size today ($N_t$) plus whatever growth I get. This growth depends on the intrinsic population growth rate $r$ and the density-dependent term, which balances the population size today with the carrying capacity $K$

## How NOT to simulate population growth

:::: {.columns}

::: {.column width='50%'}

```{r}
#| code-fold: false
n_0 <- 10
n_1 <- n_0 + (1.05 * n_0 * (1 - (n_0 / 100)))
n_2 <- n_1 + (1.05 * n_1 * (1 - (n_1 / 100)))
n_3 <- n_2 + (1.05 * n_2 * (1 - (n_2 / 100)))
n_4 <- n_3 + (1.05 * n_3 * (1 - (n_3 / 100)))
n_5 <- n_4 + (1.05 * n_4 * (1 - (n_4 / 100)))
n_6 <- n_5 + (1.05 * n_5 * (1 - (n_5 / 100)))
n_7 <- n_6 + (1.05 * n_6 * (1 - (n_6 / 100)))
n_8 <- n_7 + (1.05 * n_7 * (1 - (n_7 / 100)))
n_9 <- n_8 + (1.05 * n_8 * (1 - (n_8 / 100)))
n_10 <- n_9 + (1.05 * n_9 * (1 - (n_9 / 100)))
```

:::

::: {.column width='50%'}

```{r}
#| echo: false
tibble(t = 0:10,
       N = c(n_0, n_1, n_2, n_3, n_4, n_5, n_6, n_7, n_8, n_9, n_10)) |> 
  ggplot(aes(x = t,
             y = N)) + 
  geom_line() +
  geom_point() +
  my_theme(text_size = 15)
```


:::

::::

- Step 1: What are we repeating?
- Step 2: Within that code, what is constant and what changes?

## Build a Function for Population Growth

```{r}
#| code-fold: false
# Function name
pop_grow <- function(N, r = 1.05, K = 100) { # Function arguments
  N_next <- N + (r * N * (1 - (N / K))) # Function body
  return(N_next) # This is a formality, but highly recommended
}
```

. . .

This allows me to do:

```{r}
#| code-fold: false
n_0 <- 10
n_1 <- pop_grow(N = n_0)
n_2 <- pop_grow(N = n_1)
n_3 <- pop_grow(N = n_2)
n_4 <- pop_grow(N = n_3)
n_5 <- pop_grow(N = n_4)
n_6 <- pop_grow(N = n_5)
n_7 <- pop_grow(N = n_6)
n_8 <- pop_grow(N = n_7)
# and so forth...
```

- This is still not good enough
- What If I have to simulate my population for 50 years?
- Do I type 50 lines of code?



<!-- ## Ideas for functions -->

<!-- 1. Load WCPFC or IATTC data and fix coordinates -->
<!-- 2. Fix EU RNPA? -->

<!-- - Themes in your ggplots -->
<!-- - Your own color palette -->
<!-- -  -->

# For loops

## For loops

- Allow us to automatically repeat operations without having to write the same code multiple times

. . .

The basic structure is this:

```{r}
#| eval: false
#| code-fold: false
for (index in list_of_possible_values_for_index) {
  Do something with my index
}
```

## For loops

```{r}
#| code-fold: false
N <- 10 # Start with a population of 10
for (i in 2:50) { # For every value of i, where i can be 0, 1, 2, 3, 4... 50
  N <- pop_grow(N = N) # Overwrite my population
  print(N) # Print the pop size to the console
}
```

## For loops, leveraging the index

```{r}
#| code-fold: false
N <- numeric(50)
N[1] <- 10 # Start with a population of 10
for (i in 1:49) { # For every value of i, where i can be 0, 1, 2, 3, 4... 50
  N[i+1] <- pop_grow(N = N[i]) # Overwrite my population
}
print(N) # Print the pop size to the console
```



# References

