---
title: "Git and GitHub"
lightbox: true
format:
  html:
    toc: true
  pdf:
    toc: true
---

# Hands-on: Building Your First Repository

The overall flow:

0. Create a GitHub account (if you don't have one), install git, authenticate ([instructions](https://jcvdav.github.io/EVR_628/docs/02_installing_git.html))
1. Create a new repository on GitHub
2. Clone it to your computer
3. Add your project files
4. Make your first commit
5. Push to GitHub

## Step 0: Check set up

1. Make sure you have git. In a terminal (Mac) or Git bash (Windows) run:

```{bash}
which git
```

2. Make sure your have introduced yourself to git. To check for your configuration, use this in a terminal:

```{bash}
git config -l
```

You should see your `user.name` and `user.email` listed, typically at the bottom.

## Step 1: Create a GitHub Repository

1. Go to [github.com](https://github.com) and sign in
2. Click the "`+`" icon in the top right
3. Select "_New repository_"
4. Choose a simple name: `EVR_628`
5. Add a description: "_Learning how to use GitHub_"
6. Make it **Public**
7. Check "_Add a README file_"
8. Select gitignore tailored to R
9. Click "_Create repository_"

## Step 2: Clone to Your Computer

**In RStudio**:

1. Top-right corner, then "New Project"
2. Choose "Version Control"
3. Select "Git"
4. Paste your repository URL
5. Choose where to save it locally
6. Click "Create Project"
7. Verify that you now have a `README.md` and `.gitignore` files

\clearpage
## Step 3: Set Up Your Project Structure

To create these folders in your repository:

```
environmental-data-project/
├── data/
│   ├── raw/
│   ├── processed/
|   └── output/
├── scripts/
│   ├── 01_processing/
│   ├── 02_analysis/
│   └── 03_content/
├── results/
│   ├── img/
│   └── tab/
├── docs/
└── README.md
```

Use:

```{r}
#| eval: false
#| code-fold: false
EVR628tools::create_dirs()
```


:::{.callout-tip}
## The _namespace_
When you only want to use a function from a package, we can avoid loading the library with `library(package_name)` and instead directly access the function via it's namespace. Here, we say `package_name::function_name()`.
:::


## Step 4: Create / Modify the Global and Project .gitignore Files

1. We will use the `usethis::git_vaccinate()`
2. Look at the documentation using `?usethis::git_vaccinate()`
3. Call `usethis::git_vaccinate()`


:::{.callout-tip}
You can always add specific files or folders by editing the `.gitignore` file or using the git pane
:::

\clearpage
## Step 5: Add Your First "content"

- Create a simple R script in `scripts/content/my_first_plot.R`
- Copy the code below and paste it in the script

```{r}
#| results: false
#| code-fold: false
#| message: true
#| error: true
# Load packages
library(EVR628tools)
library(tidyverse)

# Load data
data(data_lionfish)

# Create a simple plot
p <- ggplot(data_lionfish,
            aes(x = total_length_mm, y = total_weight_gr)) +
  geom_point()

p

# Save plot
# The above shouldn't work, why?
ggsave(plot = p, filename = "results/figures/first_plot.png")
```

## Step 6: Update Your README

Create a comprehensive README.md

```markdown
# My repo for EVR 628

## Author

[Your Name and email? GitHub username?]

## Description

Analysis of environmental data for EVR 628 course.

## Project Structure

- `scripts/`: R scripts for analysis
- `results/`: Output figures and tables

```

## Step 7: Stage-Commit-Push

1. Stage and then commit your R script
2. Push
3. Stage and then commit your README
4. Stage and then commit your plot
5. Push

# Now Let's Collaborate

## 1) Simple collaboration

- Find a partner
- Decide who will be partner A and who will be partner B

:::: {.columns}

::: {.column width='50%'}
### Partner A

1. In your repo's page on GitHub, go to settings -> collaborators -> Add collaborator
2. Add partner B's username and send an invitation
3. Wait until B tells you they are done.

... wait until partner B tells you...

4. Pull
:::

::: {.column width='50%'}
### Partner B

1. Give partner A your username
2. Check your email and accept the invite
3. Go to the repository
4. Repeat the cloning process:
  - You will have a new project
  - Use a different location (I suggest desktop so you can delete it once you are done)
5. Make one change to partner A's figure, and update the exported image (e.g. `xlab(Total Length (mm))` or `color = "blue"`...)
6. Stage the files, then commit, and push
7. Tell partner A you are done
:::

::::

Changes to make by B:

```{r}
#| code-fold: false
#| code-line-numbers: "6"
# Example: environmental_data_analysis.R
library(EVR628tools)
library(tidyverse)

data(data_lionfish)

# Create a simple plot
p <- ggplot(data_lionfish,
            aes(x = total_length_mm, y = total_weight_gr)) +
  geom_point(color = "blue") +  # <--- This is one change
  xlab("Total Length (cm)")     # <--- This is another change

p
```


- Now switch who is A and who is B and repeat



