---
title: "03_git_and_github_live"
lightbox: true
format:
  html:
    toc: true
  pdf:
    documentclass: article
---

# Hands-on: Building Your First Repository

## Hands-on: Building Your First Repository

**Let's create a repository for your environmental data project!**

1. Create a GitHub account (if you don't have one)
2. Create a new repository on GitHub
3. Clone it to your computer
4. Add your project files
5. Make your first commit
6. Push to GitHub

## Step 1: Create a GitHub Repository

1. Go to [github.com](https://github.com) and sign in
2. Click the "+" icon in the top right
3. Select "New repository"
4. Choose a descriptive name: `environmental-data-project`
5. Add a description: "Analysis of environmental data for EVR 628"
6. Make it **Public** (for portfolio building)
7. Check "Add a README file"
8. Click "Create repository"

## Step 2: Clone to Your Computer

**In RStudio**:
1. Go to File → New Project
2. Choose "Version Control"
3. Select "Git"
4. Paste your repository URL
5. Choose where to save it locally
6. Click "Create Project"

## Step 3: Set Up Your Project Structure

Create these folders in your repository:

```
environmental-data-project/
├── data/
│   ├── raw/
│   ├── processed/
|   └── output/
├── scripts/
│   ├── 01_processing/
│   ├── 02_analysis/
│   └── 03_content/
├── results/
│   ├── figures/
│   └── tables/
├── docs/
└── README.md
```

Use:

```{r}
#| eval: false
#| code-fold: false
EVR628tools::create_dirs()
```


## Step 4: Create a .gitignore File

Create a `.gitignore` file to exclude unnecessary files:

```gitignore
# R specific
.Rhistory
.RData
.Ruserdata
*.Rproj.user/

# Data files (if large)
*.csv
*.rds
*.xlsx

# Results (if large)
*.png
*.pdf
*.jpg

# System files
.DS_Store
Thumbs.db
```

**Note**: You can always add specific files with `git add -f filename`, edit the `.gitignore` file or use the git pane

## Step 5: Add Your First Analysis

Create a simple R script in `scripts/analysis/`:
```{r}
#| eval: false
#| code-fold: false
# Example: environmental_data_analysis.R
library(EVR628tools)
library(tidyverse)

# Create a simple plot
p <- ggplot(data_milton, aes(x = iso_time, y = wind_speed)) +
  geom_point()

# Save plot
ggsave(plot = p, filename = "results/figures/first_plot.png") # <-- This shouldn't work, why?
```


Commit and push:

## Step 6: Update Your README

Create a comprehensive README.md

```markdown
# Environmental Data Project

## Description
Analysis of environmental data for EVR 628 course.

## Project Structure
- `data/`: Raw and processed data files
- `scripts/`: R scripts for analysis
- `results/`: Output figures and tables
- `docs/`: Documentation

## Getting Started
1. Clone this repository
2. Install required R packages
3. Run analysis scripts

## Author
[Your Name]
```

Commit and push your README file





