---
title: "Coding Principles Exercises"
lightbox: true
format:
  html:
    toc: true
  pdf:
    toc: true
---

```{r}
#| echo: false
#| warnings: false
#| message: false
library(EVR628tools)
library(tidyverse)
```

# Exercise 1: Object Classes and Assignment

## Part A: Set up

1. Open R Studio and navigate to your EVR628 project
2. Set up a code snippet for a header

```r
################################################################################
# title
################################################################################
#
# Your Name Here
# Your email here
# date
#
# Description
#
################################################################################
```

3. Create a new R script and save it as `objects_and_classes.R`
4. Use your new code snippet to add the header

## Part B: Create and Check Object Classes

In your R Script, create objects of different classes and check their types:

1. A character object called `my_name` with your name
2. A numeric object called `my_lucky_number` with your lucky number (or any number, it doesn't have to be your lucky number)
3. A logical object called `is_student` with `TRUE` or `FALSE` (note the all caps)
4. Check the class of each object using `class()`

```{r}
my_name <- "JC" # character object called `my_name` with your name
my_lucky_number <- 32 # numeric object called `my_lucky_number` with your lucky number
is_student <- FALSE # logical object called `is_student` with TRUE or FALSE

# Check the class of each object using `class()`
class(my_name)
class(my_lucky_number)
class(is_student)
```

## Part C: Object Coercion

Let's see what happens when you try converting objects. In the R script, write and then execute code that will: 

1. Convert your lucky number to `character` using `as.character(my_lucky_number)`
2. Convert `TRUE` to numeric using `as.numeric()`
3. Try to convert `"hello"` to `numeric` - what happens?
4. Advanced: Use `|>` to build a code pipeline to:
    a. Take your lucky number
    b. Convert it to character
    c. Get it's class

```{r}
#| code-fold: true
# Note that I am overwriting the object. If I had only done
# `as.character(my_lucky_number)`, then the coerced output would have been
# printed to the console.
my_lucky_number <- as.character(my_lucky_number) # Convert your lucky number to `character`

# Check that it worked
class(my_lucky_number)
as.numeric(TRUE) # Convert `TRUE` to numeric using `as.numeric()`
as.numeric("hello") # Try to convert `"hello"` to `numeric` - what happens?

# Build a pipeline
my_lucky_number <- 32 # Start with my_lucky_number as numeric again

# And now build the pipeline
my_lucky_number |> 
  as.character() |> 
  class()
```

# Exercise 2: Vectors and Operations

## Part A: Create and Manipulate Vectors

1. Clean you environment (use the broom icon)
2. Create a numeric vector called `length_m` with values: 6, 4.1, 2.8, 5.5, 3.9, 5.8
3. Create a character vector called `shark_species` with: Great White Shark, Lemon Shark, Bull Shark, Hammerhead Shark, Mako Shark, and Great White Shark (yes, white shark again)
4. How many _variables_ do you have in your environment?
5. How many length _observations_ do we have? Find the length of both vectors using `length()`
6. How many unique _species_ do we have? (Hint, use `|>` to build a pipeline)
7. Calculate the mean length of all sharks using `mean()`
8. Find the maximum length using `max()`

```{r}
#| code-fold: true
# Q2
length_m <- c(6, 4.1, 2.8, 5.5, 3.9, 5.8) 
# Q3
shark_species <- c("Great White Shark", "Lemon Shark", "Bull Shark", 
                   "Hammerhead Shark", "Mako Shark", "Great White Shark")
# Q5
length(length_m)

# Q6
shark_species |> 
  unique() |> 
  length()

# Q7
mean(x = length_m)

# Q8
max(length_m)
```

::: {.callout-important}
- When passing arguments to functions, use `=` not `<-`
- When creating objects, use `<-` not `=`
:::


## Part B: Vector Operations and Indexing

1. Extract the first 3 shark species using indexing with `[]` and save them to an object called `first_3`
2. Extract shark species where maximum length is greater than 4 meters
3. Assuming the values in `length_m` and `sharks_species` are ordered so that they match each other, find the shark species that is the largest
4. Calculate the mean length for all great white sharks

```{r}
#| code-fold: true
# Q1
first_3 <- shark_species[1:3] # Extract values 1 through 3 and assign them
first_3 # See the values I assigned

# Q2
shark_species[length_m > 4] #Extract shark species where maximum length is greater than 4 meters

# Q3
shark_species[length_m == max(length_m)]

# Q4 (Option 1, no pipes)
# Read as : "Calculate the mean of lengths where shark species matches Great White Shark
mean(length_m[shark_species == "Great White Shark"])

# Q4 (Option 2, with pipe)
length_m[shark_species == "Great White Shark"] |> # Read as: Extract lengths where name matches great white shark AND THEN ...
  mean() # Calculate the mean
```

---

Likely pause here

---

# Exercise 3: Data Frames and Tibbles

## Part A: Estimate the effect of a Marine Protected Area on Biomass

1. Start a new script called `MPA_analysis`, add a comment outline and then load the `EVR628tools` and `tidyverse` packages.

2. Load and inspect the new `?data_MPA`
    a. What are the dimensions of the data?
    b. What are the column names?
    c. How many `unique()` sites are there?
    b. How many `unique()` years?
    e. Visualize the trends in biomass through time and across sites

3. Create four objects containing:
    a. Mean biomass _inside_ the MPA **before** it was protected
    b. Mean biomass _inside_ the MPA **after** it was protected
    c. Mean biomass _outside_ the MPA **before** the MPA was created
    d. Mean biomass _outside_ the MPA **after** the MPA was created

Hint: Use a combination of subsetting (`[ ]`), relational (`==`), and logical operators (`&`)


4. Then, calculate:
    a. Change after vs before for the protected site
    b. Change after vs before for the unprotected site

5. Finally, calculate the difference between these two values. This is called the *naive* difference-in-differences estimate. You are calculating the differences across treatments and across time. See @villasenor2018user and @lynham2024evidence for details.

6. You now have four values, and you now where they come from. Assemble them into a new data.frame that
allows you to build a figure where the x-axis is time, the y-axis is mean biomass, and colors are given by the protection status of the site.

7. How would we go about adding standard deviations?


```{r}
#| code-fold: true
library(EVR628tools)
library(tidyverse)
data("data_MPA")

# 2e. Visualize
ggplot(data_MPA, aes(x = time, y = biomass, color = id)) + 
  geom_line() +
  geom_point()
```

A better visualization:

```{r}
#| code-fold: true
ggplot(data_MPA, aes(x = time, y = biomass, color = protected == 1)) + 
  geom_point() +
  geom_smooth()
```

```{r}
#| code-fold: true
## Manually calculate DiD ------------------------------------------------------
# 3a.
mean_protected_before <- mean(data_MPA$biomass[data_MPA$protected == 1 & data_MPA$after == 0])

# 3b.
mean_protected_after <- mean(data_MPA$biomass[data_MPA$protected == 1 & data_MPA$after == 1])

# 3c.
mean_control_before <- mean(data_MPA$biomass[data_MPA$protected == 0 & data_MPA$after == 0])

# 3d.
mean_control_after <- mean(data_MPA$biomass[data_MPA$protected == 0 & data_MPA$after == 1])
```

```{r}
#| code-fold: true
# Difference in time for each site
# 4a.
dif_protected <- mean_protected_after - mean_protected_before

# 4b.
dif_control <- mean_control_after - mean_control_before

# Differences across sites
dif_in_dif <- dif_protected - dif_control

dif_in_dif
```

```{r}
#| code-fold: true
# 6. Assemble the four values into a data.frame. Each vector I pass becomes a column,
# so I have to keep the order consistent across all three of them.
summary_data <- data.frame(period = c("before", "after", "before", "after"),
                           status = c("reserve", "reserve", "control", "control"),
                           mean_biomass = c(mean_protected_before,
                                            mean_protected_after,
                                            mean_control_before,
                                            mean_control_after))

summary_data

# Now plot it
ggplot(summary_data, aes(x = period, y = mean_biomass, fill = status)) +
  geom_col(position = "dodge", color = "black")
```

```{r}
#| code-fold: true
# 7. Exactly the same subsetting as before, but with sd() instead of mean()
sd_protected_before <- sd(data_MPA$biomass[data_MPA$protected == 1 & data_MPA$after == 0])
sd_protected_after <- sd(data_MPA$biomass[data_MPA$protected == 1 & data_MPA$after == 1])
sd_control_before <- sd(data_MPA$biomass[data_MPA$protected == 0 & data_MPA$after == 0])
sd_control_after <- sd(data_MPA$biomass[data_MPA$protected == 0 & data_MPA$after == 1])

# Add them as a new column, in the same order as the means above
summary_data$sd_biomass <- c(sd_protected_before,
                             sd_protected_after,
                             sd_control_before,
                             sd_control_after)

summary_data

# And show them as error bars
ggplot(summary_data, aes(x = period, y = mean_biomass, fill = status)) +
  geom_col(position = "dodge", color = "black") +
  geom_errorbar(aes(ymin = mean_biomass - sd_biomass,
                    ymax = mean_biomass + sd_biomass),
                width = 0.1,
                position = position_dodge(width = 1))
```

Extra: We can also do this with a linear model

```{r}
#| code-fold: true
# Can also do this as a linear model
lm(biomass ~ after + protected + after * protected,
      data = data_MPA) |> 
  summary()
```


\clearpage

# Exercise 4: Code Style and Documentation

## Part A: Fix Code Style

Fix the style issues in this code:

```r
mydataframe=data.frame(species=c("Great White","Tiger", "Bull"),
length=c(4.5,3.2, NA))
mean(mydataframe$length,na.rm=TRUE)
```

```{r}
# My improved code will be here

```

Note: What's with than `na.rm = TRUE`?

## Part B: Add Comments and Section Headers

1. Add meaningful comments to your R script
