---
title: "A presentation about FL hurricanes"
subtitle: "A brief analysis"
author: "Juan Carlos Villaseñor-Derbez (JC)"
institute: "Rosenstiel School of Marine, Atmospheric, and Earth Science and Institute for Data Science and Computing"
format:
  html:
    echo: true
    code-fold: true
---

# Intro + Objectives

- Hurricanes affect the FL peninsula almost every year 
- Our goal is to identify the county that has been hit the most frequently (2022-2024)

# Methods

- Combine hurricane track data (vector) and county data (vector)
- Count number intersections per county
- Find the most hit county

```{r}
# Load packages
library(tidyverse)
library(EVR628tools)
library(here)
library(sf)
library(mapview)
# Load data
data("data_hurricanes")
FL_counties <- read_sf(here("data/raw/Florida_County_Boundaries_with_FDOT_Districts_-801746881308969010.gpkg")) |> 
  janitor::clean_names() |> 
  st_set_geometry("geometry") |> 
  select(county_name = name,
         county_fips = first_fips)
# Find n storms per county
county_hur <- st_join(FL_counties, data_hurricanes) |> 
  group_by(county_name, county_fips) |> 
  summarize(n_storms = n_distinct(name,
                                  na.rm = T),
            .groups = "drop")
```

# Results

- The county with the most hurricanes is **`r county_hur$county_name[which(county_hur$n_storms == max(county_hur$n_storms))]`** county
- Between 2022 and 2024, it was hit `r max(county_hur$n_storms)` times

# Results

```{r}
FL_hurricanes <- st_filter(data_hurricanes, FL_counties) |> 
  st_crop(FL_counties)

most_hit_county <- county_hur |> 
  filter(n_storms == max(n_storms))

mapview(county_hur,
        zcol = "n_storms",
        layer.name = "# Storms") + 
  mapview(most_hit_county,
          col.regions = "transparent",
          color = "red",
          lwd = 4,
          layer.name = "Taylor county") +
  mapview(FL_hurricanes,
          zcol = "name",
          layer.name = "Storm name",
          color = palette_UM(7))
```

# Final remarks

- Using Quarto markdown allows you to include R-generated code in cool slides

- How I would probably go about building these slides

```{mermaid}
flowchart LR
  A(data_hurricanes) --> C{processing/hurricanes_per_county.R}
  B(FL counties) --> C
  C --> D(county_hur.gpkg)
  C --> E(FL_hurricanes.gpkg)
  D --> F[slides.qmd
  Build leaflet in code chunk]
  E --> F
```

